No one-size-fits-all solution to clean GBIF

### 2020-02-29 manuscript_peerj.tex
peerj format

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={No one-size-fits-all solution to clean GBIF},
  pdfauthor={Alexander Zizka1,2,; Fernanda Antunes Carvalho4,; Alice Calvente4,; Ana Paula Alves da Silva4,; Andressa Cabral5,; Arthur de Souza Soares4,; Eduardo Calisto Tomaz4,; Fernanda Gondim Lambert Moreira4,; Filipe Cabreirinha Serrano6,; Jéssica Fernanda Ramos Coelho4,; Mabel Rocio Baez-Lizarazo,7; Mariana Ramos Fantinati4,; Matheus Colli-Silva5,; Moabe Ferreira Fernandes8,; Nathália Michelly da Cunha Santos4,; Paolla Gabryelle Cavalcante de Souza4,; Renata Clicia dos Santos‐Costa4,; Thais Ferreira-Araújo4,; Tiago Andrade Borges Santos9,; Tiago Luiz Vieira9,; Valéria Fonseca Vale4,; Alexandre Antonelli10,11,12},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{caption}
\usepackage{fancyhdr}
\usepackage{booktabs}
\pagestyle{fancy}
\fancyhead[L]{Automated filter effect}
\fancyhead[LE,RO]{\thepage}
\fancyfoot{}
\usepackage{float}
\floatplacement{figure}{H}
\usepackage{lineno}
\linenumbers
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{No one-size-fits-all solution to clean GBIF}
\author{Alexander Zizka\textsuperscript{1,2}, \and Fernanda Antunes Carvalho\textsuperscript{4}, \and Alice Calvente\textsuperscript{4}, \and Ana Paula Alves da Silva\textsuperscript{4}, \and Andressa Cabral\textsuperscript{5}, \and Arthur de Souza Soares\textsuperscript{4}, \and Eduardo Calisto Tomaz\textsuperscript{4}, \and Fernanda Gondim Lambert Moreira\textsuperscript{4}, \and Filipe Cabreirinha Serrano\textsuperscript{6}, \and Jéssica Fernanda Ramos Coelho\textsuperscript{4}, \and Mabel Rocio Baez-Lizarazo,\textsuperscript{7} \and Mariana Ramos Fantinati\textsuperscript{4}, \and Matheus Colli-Silva\textsuperscript{5}, \and Moabe Ferreira Fernandes\textsuperscript{8}, \and Nathália Michelly da Cunha Santos\textsuperscript{4}, \and Paolla Gabryelle Cavalcante de Souza\textsuperscript{4}, \and Renata Clicia dos Santos‐Costa\textsuperscript{4}, \and Thais Ferreira-Araújo\textsuperscript{4}, \and Tiago Andrade Borges Santos\textsuperscript{9}, \and Tiago Luiz Vieira\textsuperscript{9}, \and Valéria Fonseca Vale\textsuperscript{4}, \and Alexandre Antonelli\textsuperscript{10,11,12}}
\date{}

\begin{document}
\maketitle

\setstretch{1.3}
\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany
\item
  Naturalis Biodiversity Center, Leiden, The Netherlands
\item
  Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil
\item
  Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil
\item
  Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil
\item
  Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil
\item
  Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil
\item
  Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil
\item
  Universidade Estadual de Feira de Santana, Bahia, Brazil
\item
  Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden
\item
  Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
\item
  Royal Botanic Gardens Kew, Richmond, United Kingdom
\end{enumerate}

\newpage{}

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF. We find that 29-67\% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect. While many flagged records might not be \emph{de facto} erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.

\hypertarget{keywords}{%
\section{Keywords}\label{keywords}}

Automated cleaning, automated conservation assessment, data quality, GBIF, Neotropics.

\newpage{}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

@@

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al., 2016). Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al., 2015; Yesson et al., 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al., 2019).

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers (e.g., Chamberlain, 2016; Jin \& Yang, 2020; Zizka, Silvestro, et al., 2019) or on additional data, such as environmental variables (Robertson, Visser, \& Hui, 2016). Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al., 2019).

Erroneous records are especially problematic in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Bachman et al., 2011; Dauby et al., 2017; Pelletier et al., 2018). These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B (``Geographic range''). These methods have been used topropose preliminary global (Stévart et al., 2019; Zizka, Azevedo, et al., 2019) and regional (Cosiaux et al., 2018; Schmidt et al., 2017) Red List assessments. However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates \textbf{(REFERENCE)}. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

@@

\hypertarget{material-and-methods}{%
\section{Material and Methods}\label{material-and-methods}}

\hypertarget{choice-of-study-taxa}{%
\subsection{Choice of study taxa}\label{choice-of-study-taxa}}

@@

\hypertarget{species-occurrence-data}{%
\subsection{Species occurrence data}\label{species-occurrence-data}}

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain, 2017) in R (GBIF.org, 2019a-p). We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone, 2014). The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.

\hypertarget{automated-cleaning}{%
\subsection{Automated cleaning}\label{automated-cleaning}}

We followed the cleaning pipeline outlined by Zizka, Silvestro, et al. (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records). In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al., 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter). These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

\hypertarget{downstream-analyses}{%
\subsection{Downstream analyses}\label{downstream-analyses}}

We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al., 2016). We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al., 2017). \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee, 2017), they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain, 2018).

\hypertarget{evaluation-of-results}{%
\subsection{Evaluation of results}\label{evaluation-of-results}}

@@

We did all analyses in the R computing environment (R Core Team, 2019), using standard libraries for data handling and visualization (Garnier, 2018; Hijmans, 2019; Ooms, 2014, 2019; Wickham, 2018). All scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.

\hypertarget{results}{%
\section{Results}\label{results}}

We retrieved a total of 215,262 species occurrence records, with a median of 3,460 records per study group and 11 records per species (Table 3, Appendix 2). We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51). The species with most records was \emph{Harengula jaguana} (19,878).

Our automated tests flagged a total of 88,648 records (Fig. \ref{fig:total}), with a median of 40.3\% per group. Overall, the most important test was for duplicated records (on average 32.9\% per taxonomic group). The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (2.1\%) and the collection year (3.3\%). The most important automated tests were for urban area (5\%), the capital coordinates (1.1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers). Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Automated flagging based on meta-data and automatic tests flagged on average 40.3\% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

@@

The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig. \ref{fig:speciesrichness}, Supplementary material 1). The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al., 2015). Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions. These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.

The accuracy of the automated conservation assessment was in the same range as found by previous studies (Nic Lughadha et al., 2019; Zizka, Azevedo, et al., 2019). The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies (Gueta \& Carmel, 2016). The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering. This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF. Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available. The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al., 2011).

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

@@

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

@@

\hypertarget{supplementary-material}{%
\section{Supplementary material}\label{supplementary-material}}

Appendix 1 - Analyses scripts

Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test

Appendix 3 - The full results of the conservation assessment

Appendix 4 - Taxon specific richness maps and comments.

\newpage{}

\hypertarget{tables}{%
\section{Tables}\label{tables}}

@@

\newpage{}

\hypertarget{figures}{%
\section{Figures}\label{figures}}

@@

\newpage{}

>>>>

%% Submissions for peer-review must enable line-numbering
%% using the lineno option in the \documentclass command.
%%
%% Preprints and camera-ready submissions do not need
%% line numbers, and should have this option removed.
%%
%% Please note that the line numbering option requires
%% version 1.1 or newer of the wlpeerj.cls file, and
%% the corresponding author info requires v1.2

\documentclass[fleqn,10pt,lineno]{wlpeerj} % for journal submissions

% ZNK -- Adding headers for pandoc

\setlength{\emergencystretch}{3em}
\providecommand{\tightlist}{
\setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{lipsum}
\usepackage[unicode=true]{hyperref}
\usepackage{longtable}

% Pandoc syntax highlighting
% See https://github.com/rstudio/rticles/issues/182

% Pandoc Header
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\title{No one-size-fits-all solution to clean GBIF}

\author[1,2]{Alexander Zizka}

\corrauthor[1,2]{Alexander Zizka}{\href{mailto:alexander.zizka@idiv.de}{\nolinkurl{alexander.zizka@idiv.de}}}
\author[3]{Fernanda Antunes Carvalho}

\author[4]{Alice Calvente}

\author[4]{Ana Paula Alves da Silva}

\author[5]{Andressa Cabral}

\author[4]{Arthur de Souza Soares}

\author[4]{Eduardo Calisto Tomaz}

\author[4]{Fernanda Gondim Lambert Moreira}

\author[6]{Filipe Cabreirinha Serrano}

\author[4]{Jéssica Fernanda Ramos Coelho}

\author[7]{Mabel Rocio Baez-Lizarazo}

\author[4]{Mariana Ramos Fantinati}

\author[5]{Matheus Colli-Silva}

\author[8]{Moabe Ferreira Fernandes}

\author[4]{Nathália Michelly da Cunha Santos}

\author[4]{Paolla Gabryelle Cavalcante de Souza}

\author[4]{Renata Clicia dos Santos‐Costa}

\author[4]{Thais Ferreira-Araújo}

\author[9]{Tiago Andrade Borges Santos}

\author[9]{Tiago Luiz Vieira}

\author[4]{Valéria Fonseca Vale}

\author[10,11,12]{Alexandre Antonelli}

\affil[1]{sDiv, German Center for Integrative Biodiversity Research (iDiv) Halle-Jena-Leipzig, Leipzig, Germany}
\affil[2]{Naturalis Biodiversity Center, Leiden, The Netherlands}
\affil[3]{Departamento de Genética, Ecologia e Evolução, Universidade Federal de Minas Gerais, Belo Horizonte, Brazil}
\affil[4]{Departamento de Botânica e Zoologia, Universidade Federal do Rio Grande do Norte, Natal, Brazil}
\affil[5]{Departamento de Botânica, Universidade de São Paulo, São Paulo, Brazil}
\affil[6]{Departamento de Ecologia, Universidade de São Paulo, São Paulo, Brazil}
\affil[7]{Universidade Federal do Rio Grande do Sul, Porto Alegre, Brazil}
\affil[8]{Departamento de Ciências Biológicas, Universidade Estadual de Feira de Santana, Feira de Santana, Brazil}
\affil[9]{Universidade Estadual de Feira de Santana, Bahia, Brazil}
\affil[10]{Gothenburg Global Biodiversity Centre, University of Gothenburg, Gothenburg, Sweden}
\affil[11]{Department for Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden}
\affil[12]{Royal Botanic Gardens Kew, Richmond, United Kingdom}

%
% \author[1]{First Author}
% \author[2]{Second Author}
% \affil[1]{Address of first author}
% \affil[2]{Address of second author}
% \corrauthor[1]{First Author}{f.author@email.com}

%

\begin{abstract}
Species occurrence records provide the basis for many biodiversity studies. They derive from geo-referenced specimens deposited in natural history collections and visual observations, such as those obtained through various mobile applications. Given the rapid increase in availability of such data, the control of quality and accuracy constitutes a particular concern. Automatic flagging and filtering are a scalable and reproducible means to identify potentially problematic records in datasets from public databases such as the Global Biodiversity Information Facility (GBIF; www.gbif.org). However, it is unclear how much data may be lost by filtering, whether the same tests should be applied across all taxonomic groups, and what is the effect of filtering for common downstream analyses. Here, we evaluate the effect of 13 recently proposed filters on the inference of species richness patterns and automated conservation assessments for 16 Neotropical taxa including plants and animals, terrestrial and marine, downloaded from GBIF. We find that 29-67\% of the records are potentially erroneous, with large variation across taxonomic groups. Tests for duplicated information, collection year, basis of record as well as urban areas and political centroids have the greatest effect. While many flagged records might not be \emph{de facto} erroneous, they could be overly imprecise and increase uncertainty in downstream analyses. Automated flagging can help in identifying problematic records, but requires customization of which tests and thresholds should be applied to the taxonomic group and geographic area under focus. Our results stress the importance of thorough exploration of the meta-data associated with species records for biodiversity research.
% Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text. Dummy abstract text.
\end{abstract}

\begin{document}

\flushbottom
\maketitle
\thispagestyle{empty}

\hypertarget{introduction}{%
\section*{Introduction}\label{introduction}}
\addcontentsline{toc}{section}{Introduction}

@@

A central challenge to the use of these publicly available species occurrence data in research are erroneous geographic coordinates (Anderson et al. 2016). Errors mostly arise because public databases integrate records collected with different methodologies in different places, at different times; often without centralized curation and only rudimentary meta-data. For instance, erroneous coordinates caused by data-entry errors or automated geo-referencing from vague locality descriptions are common (Maldonado et al. 2015; Yesson et al. 2007) and cause recurrent problems such as records of terrestrial species in the sea, records with coordinates assigned to the centroids of political entities, or records of species in captivity (Zizka, Silvestro, et al. 2019).

Manual data cleaning based on expert knowledge can detect these issues, but it is only feasible on small taxonomic or geographic scales, and it is time-consuming and difficult to reproduce. As an alternative, automated flagging methods to identify potentially problematic records have been proposed as a scalable option, as they are able to deal with datasets containing up to millions of records and many different taxa. Those methods are usually based on geographic gazetteers (e.g., Chamberlain 2016; Zizka, Silvestro, et al. 2019; Jin and Yang 2020) or on additional data, such as environmental variables (Robertson, Visser, and Hui 2016). Additionally, filtering procedures based on record meta-data, such as collection year, record type, and coordinate precisions have been proposed to improve the quality of publicly available occurrence records for biodiveristy research (Zizka, Silvestro, et al. 2019).

Erroneous records are especially problematic in conservation, where stakes are high. Recently proposed methods for automated conservation assessments could support the formal assessment procedures for the global Red List of the International Union for the Conservation of Nature (IUCN) (Dauby et al. 2017; Bachman et al. 2011; Pelletier et al. 2018). These methods approximate species' range size, namely the Extent of Occurrence (EOO, which is the area of a convex hull polygon comprising all records of a species), the Area of Occupancy (AOO, which is the sum of the area actually occupied by a species, calculated based on a small-scale regular grid), and the number of locations for a preliminary conservation assessment following IUCN Criterion B (``Geographic range''). These methods have been used topropose preliminary global (Stévart et al. 2019; Zizka, Azevedo, et al. 2019) and regional (Schmidt et al. 2017; Cosiaux et al. 2018) Red List assessments. However, all metrics but especially EOO, are sensitive to individual records with erroneous coordinates \textbf{(REFERENCE)}. Automated conservation assessments may therefore be biased, particularly if the number of records are few, as it is the case for many tropical species.

@@

\hypertarget{material-and-methods}{%
\section*{Material and Methods}\label{material-and-methods}}
\addcontentsline{toc}{section}{Material and Methods}

\hypertarget{choice-of-study-taxa}{%
\subsection*{Choice of study taxa}\label{choice-of-study-taxa}}
\addcontentsline{toc}{subsection}{Choice of study taxa}

@@

\hypertarget{species-occurrence-data}{%
\subsection*{Species occurrence data}\label{species-occurrence-data}}
\addcontentsline{toc}{subsection}{Species occurrence data}

We downloaded occurrence information for all study groups from www.gbif.org using the \texttt{rgbif} v1.4.0 package (Chamberlain 2017) in R (GBIF.org, 2019a-p). We downloaded only records with geographic coordinates and limited the study area to latitudes smaller than 33\(^\circ\) N and longitudes smaller than 35\(^\circ\) W and larger than 120\(^\circ\) W reflecting the Neotropics, our main area of expertise (Morrone 2014). The natural distribution of all included taxa are confined to the Neotropics except Arhynchobatidae, Diogenidae, Dipsadidae, \emph{Gaylussacia}, Iridaceae, Neanuridae, and \emph{Tillandsia} for which we only obtained the Neotropical occurrences.

\hypertarget{automated-cleaning}{%
\subsection*{Automated cleaning}\label{automated-cleaning}}
\addcontentsline{toc}{subsection}{Automated cleaning}

We followed the cleaning pipeline outlined by Zizka, Silvestro, et al. (2019) and first filtered the raw data downloaded from GBIF (``raw'', hereafter) using meta-data for those records for which they were available, removing: 1) records with a coordinate precision below 100 km (as this represent the grain size of many macro-ecological analyses); 2) fossil records and records of unknown source; 3) records collected before 1945 (before the end of the Second World War, since coordinates of old records are often imprecise); and 4) records with an individual count of less than one and more than 99. Furthermore, we rounded the geographic coordinates to four decimal precision and retained only one record per species per location (i.e.~test for duplicated records). In a second step, we used the \texttt{clean\_coordinates} function of the \texttt{CoordinateCleaner\ v2.0-11} package (Zizka, Silvestro, et al. 2019) with default options to flag errors that are common to biological data sets (``filtered'', hereafter). These including: coordinates in the sea for terrestrial taxa, coordinates containing only zeros, coordinates assigned to country and province centroids, coordinates within urban areas, and coordinates assigned to biodiversity institutions. See Table 2 for a summary of all flagging tests we used.

\hypertarget{downstream-analyses}{%
\subsection*{Downstream analyses}\label{downstream-analyses}}
\addcontentsline{toc}{subsection}{Downstream analyses}

We first generated species richness maps using 100x100 km grid cells for the raw and filtered datasets respectively, using the package \texttt{speciesgeocodeR\ v2.0-10} (Töpel et al. 2016). We then performed an automated conservation assessment for all study groups based on both datasets using the \texttt{ConR\ v1.2.4} package (Dauby et al. 2017). \texttt{ConR} estimates the EOO, AOO, and the number of locations, and then suggests a preliminary conservation status based on Criterion B of the global IUCN Red List. While these assessments are preliminary (see IUCN Standards and Petitions Subcommittee 2017), they can be a proxy used by the IUCN to speed up full assessments. We then benchmarked the preliminary conservation assessments against the global IUCN Red List assessments as far as available for the same taxa, which we obtained from www.iucn.org via the \texttt{rredlist\ v.0.5.0} package (Chamberlain 2018).

\hypertarget{evaluation-of-results}{%
\subsection*{Evaluation of results}\label{evaluation-of-results}}
\addcontentsline{toc}{subsection}{Evaluation of results}

@@

We did all analyses in the R computing environment (R Core Team 2019), using standard libraries for data handling and visualization (Wickham 2018; Garnier 2018; Ooms 2014, 2019; Hijmans 2019). All scripts are available from \url{https://github.com/idiv-biodiversity/effects_of_automated_cleaning}, or from the Appendix 1 in the electronic supplement of this study.

\hypertarget{results}{%
\section*{Results}\label{results}}
\addcontentsline{toc}{section}{Results}

We retrieved a total of 218,899 species occurrence records, with a median of 2,844.5 records per study group and 10 records per species (Table 3, Appendix 2). We obtained most records for Dipsadidae (64,249) and fewest for \emph{Thozetella} (51). The species with most records was \emph{Harengula jaguana} (19,878).

Our automated tests flagged a total of 91,434 records (Fig. \ref{fig:total}), with a median of 42.7\% per group. Overall, the most important test was for duplicated records (on average 35.5\% per taxonomic group). The filtering steps based on record meta-data that flagged the largest number of records were the basis of records (5.9\%) and the collection year (3.4\%). The most important automated tests were for urban area (8.6\%), the capital coordinates (1\%) and the centroids of political units (0.5\%, see Table 3 and Supplementary material 1 for further details and the absolute numbers). Only few records were flagged by the zero coordinates and biodiversity institution tests (Fig. \ref{fig:split}), and no records were flagged by the sea test.

@@

\hypertarget{discussion}{%
\section*{Discussion}\label{discussion}}
\addcontentsline{toc}{section}{Discussion}

Automated flagging based on meta-data and automatic tests flagged on average 42.7\% of the records per taxonomic group. The filters for basis of record, duplicates, collection year, and urban areas flagged the highest fraction of records (Question 1). The importance of different tests was similar across taxonomic groups, with few exceptions especially for the tests on basis of record, collection year, capital coordinates, and urban areas (Question 2). The results for species richness were similar between the raw and filtered data with some improvements by the filters. We found little impact of filtering on the accuracy of the automated conservation assessments (Question 3).

@@

The general agreement between the species richness maps based on raw and filtered data was encouraging, for the use of this data for large-scale biogeographic research (Fig. \ref{fig:speciesrichness}, Supplementary material 1). The filter based on political centroids had an important impact on species richness patterns, which is congruent with the results from a previous study in the coffee family (Maldonado et al. 2015). Records assigned to country or province centroids are often old records, which are geo-referenced at a later point based on vague locality descriptions. These records are at the same time more likely to represent dubious species names, since they might be old synonyms or type specimens of species that have only been collected and described once, which are erroneously increasing species numbers.

The accuracy of the automated conservation assessment was in the same range as found by previous studies (Nic Lughadha et al. 2019; Zizka, Azevedo, et al. 2019). The similar accuracy of the raw and filtered dataset for the automated conservation assessment was surprising, in particular given the EOO and AOO reduction observed in the filtered dataset (Table 4) and the impact of errors on spatial analyses observed in previous studies (Gueta and Carmel 2016). The robustness of the automated assessment was likely due to the fact that the EOO for most species was large, even after the considerable reduction caused by filtering. This might be caused by the structure of our comparison, which only included species that were evaluated by the IUCN Red List (and not considered as Data Deficient) and at the same time had occurrences recorded in GBIF. Those inclusion criteria are likely to have, biased the datasets towards species with large ranges, since generally more data for them are available. The robustness of automated conservation assessments to data quality is encouraging, although these methods are only an approximation, especially for species with few collection records (Rivers et al. 2011).

\hypertarget{conclusions}{%
\section*{Conclusions}\label{conclusions}}
\addcontentsline{toc}{section}{Conclusions}

@@

\hypertarget{acknowledgments}{%
\section*{Acknowledgments}\label{acknowledgments}}
\addcontentsline{toc}{section}{Acknowledgments}

@@

\hypertarget{supplementary-material}{%
\section*{Supplementary material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary material}

\begin{itemize}
\item
  Appendix 1 - Analyses scripts
\item
  Appendix 2 - Table S1, reporting the absolute number of flagged records per taxonomic group and test
\item
  Appendix 3 - The full results of the conservation assessment
\item
  Appendix 4 - Taxon specific richness maps and comments.
\end{itemize}

\hypertarget{tables}{%
\section*{Tables}\label{tables}}
\addcontentsline{toc}{section}{Tables}

@@

\hypertarget{figures}{%
\section*{Figures}\label{figures}}
\addcontentsline{toc}{section}{Figures}

\begin{figure}
\includegraphics[width=0.9\linewidth]{./figures_tables/Figure_study_groups} \caption{Example of study taxa. \textbf{A)} Pilosocereus pusillibaccatus, picture by Pamela Lavor; \textbf{B)} Conchocarpus macrocarpus, Juliana El-Ottra; \textbf{C)} Tillandsia recurva, EC Tomaz; \textbf{D} \textit{Oxyrhopus guibei} (Dipsadidae), Filipe C Serrano;\textbf{E)} Neanuridae (\textit{Aethiopella ricardoi}), Raiane Vital da Paz; \textbf{F)} \textit{Tocoyena formosa}, Fernanda GL Moreira; \textbf{G)} \textit{Harengula jaguana}, Thais Ferreira-Araujo; \textbf{H)} \textit{Gaylussacia decipiens}, Luiz Menini Neto; \textbf{I)}  extit{Ooocephalus foliosus}, Arthur Soares; \textbf{J)} \textit{Tityus carvalhoi}, Renata C Santos-Costa; \textbf{K)} \textit{Prostheca vespa}, Tiago veira}\label{fig:species}
\end{figure}

@@
